Tests of measurement invariance via stochastic processes: {S}ome properties

### 2013-08-08 frontiers/frontiers.Rnw
Frontiers manuscript: Added to background and tutorial sctest.lavaan: Added fnl argument in case critical values                have already been simulated.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@235 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

In the following section, we first briefly review the 
theoretical framework of proposed statistical tests. Subsequently, we study the tests'
ability through simulations that mimic practical research. 
Finally, some suggestion on the tests' use in practice is provided.

@@

As currently implemented for SEM, the statistical tests described in
this paper can be applied to models that are estimated via a
multivariate normal or Wishart likelihood (or discrepancy) function,
with extension to other discrepancy functions being straightforward.
The tests are carried out following model estimation, making use of output
associated with the fitted model.  In general, we fit a model that
restricts parameters to be equal across observations, then carry out a
posthoc test to examine whether specific parameters vary
across observations.  This procedure is similar in spirit to the
calculation of modification indices CITE and to Lagrange multiplier
tests CITE, and, in fact, those statistics can be viewed as  
special cases of the family described here.

@@

To use these scores for testing, we order individuals according to
an auxiliary variable (the variable against which we are testing 
measurement invariance) and look for trends in the
scores (roughly speaking).  For example, imagine that we are testing
for measurement 
invariance w.r.t.\ age.  If the manifest variables violate measurement
invariance here, then some parameter estimates may be too large for
young individuals and too small for old individuals (say).  This
result would be reflected in the scores, where young individuals'
scores would be greater than zero and old individuals' scores would be
less than zero.  Conversely, if measurement invariance holds, then all
individuals' scores will fluctuate around zero.

To formalize the ideas in the previous paragraph, we focus on a cumulative
sum of the ordered scores.  Specifically, we focus on how the
cumulative sum fluctuates as more individuals' scores are added to it,
starting with the youngest and ending with the oldest individual
(assuming that age is the auxiliary variable of interest).  Under the
hypothesis of measurement invariance, a central limit theorem can be
used to show that the fluctuation of the cumulative sum follows a
Brownian bridge.  This result allows us to calculate $p$-values and
critical values for test statistics under the hypothesis of
measurement invariance.  This includes test statistics associated with
all model parameters and subsets of model parameters.

%% TODO focus on ordinal
Multiple test statistics are available, depending on how one
summarizes the behavior of the cumulative sum of scores.  One could
take the absolute maximum that the cumulative sum attains for any
parameter of interest, resulting in a {\em double max} statistic.
Alternatively, one could sum the (squared) cumulative sum across parameters
of interest and take the maximum across individuals, resulting in a
{\em maximum Lagrange multiplier} statistic 
\cite<see>[for further discussion of this idea]{MerZei13}.

@@

<<theta estimate>>=
# estimate theta 
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by conducting 
log likelihood test, which is done by anova function in R. 
<<LRT>>=
# LRT
LRT<-as.matrix(anova(theta, thetaAB))
print(LRT,na.print="", quote=FALSE)
@
\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ 
is a relatively-small number of categories and the sample size 
is not too large (LRT is sensitive to large sample and will 
return "false" significance results, details see next section). 
However, when $V$ (in this case, age) continuous variable, however, 
multiple-group analysis usually cannot be performed because there are 
no existing groups. Instead, we can follow the procedure proposed by 
Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters 
are restricted to be equal across across all individuals and then examine 
how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. 
To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and 
define the $k$ dimensional cumulative score process as 
\begin{equation}
\label{equ:B}
B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}
{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ 
is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. 
~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. 
$V$ and decorrelates the scores between $k$ components of the parameters in the model. 
Further, there exists a functional central limit theorem 
that allows us to make formal inference with this cumulative 
score process. Assuming that individuals are independent and the usual 
ML regularity condition hold, it is possible to show that(addref) 
\begin{equation}
B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
\end{equation} 
where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ 
is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement 
invariance by comparing the behavior of cumulative score process to that of 
a Brownian bridge. In practice, there are three test statistics can be used: 
double maximum, CvM and maxLM. 
\begin{equation}
DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
\end{equation}
\begin{equation}
CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
\end{equation}
\begin{equation}
maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
\end{equation}

Specifically, we will use the previous example, except assuming ordinal agegroup to be continuous.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we will use the gefp function in strucchange package to 
get the $B(\hat{\theta}_{ij})$. After that, we can get the 
three test statistics' significance by using sctest function.

@@

<<Get Estimate Alternative Way>>=
DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
DM
CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
CvM
#maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
#maxLM (Problem: nonconsistent with sctest(gefp1))
@

@@

\subsection{Ordinal Test}
However, in this dataset, $V$ is actually ordinal instead of 
continuous. There is only a partial ordering of individuals 
with respect to $V$,i.e. observations with the same level 
of $V$ have no unique ordering. The ordinal statistics 
proposed here are similar to those described in Equation 
and above, except that we focus on "bins" of individuals 
at each level of the ordinal variable. That is, instead 
of aggregating over all $i=1,...n$ individuals, we 
first compute cumulative proportions $t_{l}(l=1,...,m-1)$ 
associated with the first $m-1$ levels of $V$. 
We then aggregate the cumulative scores only over 
$i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics 
related can be written as 
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ 
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} 
      ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}

The difference between the continuous test and ordinal 
test is that we don't force arbitary order on the observations 
within the same age groups. Therefore, this ordinal test should 
reflect more of the data. In terms of code, we still need the gefp 
function to get the B estimate. After that, we can use different 
functions to aggregate and scale B estimation. 
The function are called ordwmax and ol2bb1.fun, respectively. 
<<Ordinal Test>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval.ord=rep(NA,2)
pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value
pval.ord[2]<-sctest(gefp1,functional=ordL2BB(gefp1))$p.value
pval.ord
@

<<Get Estimate Alternative Way for Ordinal>>=
WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
WDMo 
maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fplot=TRUE)
maxLMo
@

@@

In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. 
Both $p-$values are clearly larger than that of the 
likelihood ratio test and neither is significant at $\alpha=0.5$

>>>>

## turn off stars
options(show.signif.stars = FALSE)

@@

In the following section, we first briefly review the 
theoretical framework of proposed statistical tests and
provide a short tutorial illustrating the use of the tests in R.
Next, we study the tests'
ability through simulations that mimic practical research. 
Finally, some suggestion on the tests' use in practice is provided.

@@

As currently implemented for SEM, the statistical tests described in
this paper can be applied to models that are estimated via a
multivariate normal or Wishart likelihood (or discrepancy) function,
with extension to other discrepancy functions being straightforward.
The tests are carried out following model estimation, making use of output
associated with the fitted model.  In general, we fit a model that
restricts parameters to be equal across observations, then carry out a
posthoc test to examine whether specific parameters appear to vary
across observations.  This procedure is similar in spirit to the
calculation of modification indices CITE and to Lagrange multiplier
tests CITE, and, in fact, those statistics can be viewed as  
special cases of the family described here.

@@

To use these scores for testing, we order individuals according to
an auxiliary variable (the variable against which we are testing 
measurement invariance) and look for trends in the
scores (roughly speaking).  For example, imagine that we are testing
for measurement 
invariance w.r.t.\ age.  If the manifest variables violate measurement
invariance here, then some parameter estimates may be too large for
young individuals and too small for old individuals (say).  This
result would be reflected in the scores, where young individuals'
scores would be greater than zero and old individuals' scores would be
less than zero (though the sign of the scores will also depend on whether a
function is being minimized or maximized).  Conversely, if measurement
invariance holds, then all 
individuals' scores will fluctuate around zero.

To formalize the ideas in the previous paragraph, we focus on a cumulative
sum of the ordered scores.  This cumulative sum may be written as
\begin{equation} \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ %\hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of $nt$ (i.e., a floor
operator) and $x_{(i)}$ reflects the individual with the
$i^{\text{th}}$-smallest value of the auxiliary variable.
We specifically focus on how the
cumulative sum fluctuates as more individuals' scores are added to it,
starting with the youngest and ending with the oldest individual
(assuming that age is the auxiliary variable of interest).  
The summation is also typically premultiplied by the inverse square
root of the information matrix, which serves to decorrelate the
fluctuation processes associated with individual model parameters.

Under the
hypothesis of measurement invariance, a central limit theorem can be
used to show that the fluctuation of the above cumulative sum follows a
Brownian bridge.  This result allows us to calculate $p$-values and
critical values for test statistics under the hypothesis of
measurement invariance.  We can obtain test statistics associated with
all model parameters, as well as test statistics associated with
subsets of model parameters.

%% TODO focus on ordinal
Multiple test statistics are available, depending on how one
summarizes the behavior of the cumulative sum of scores.  One could
take the absolute maximum that the cumulative sum attains for any
parameter of interest, resulting in a {\em double max} statistic.
Alternatively, one could sum the (squared) cumulative sum across parameters
of interest and take the maximum across individuals, resulting in a
{\em maximum Lagrange multiplier} statistic 
\cite<see>[for further discussion of this idea]{MerZei13}.
%% TODO equations of statistics?

Importantly, these statistics were derived for situations where
individuals are uniquely ordered according to the auxiliary variable.
This is not always the case for measurement invariance applications,
where the auxiliary variable is often ordinal.  To remedy this
situation, \citeA{MerFanZei} extended the framework to situations
where one has an ordinal auxiliary variable of interest.  Essentially,
one allows all individuals with the same value of the auxiliary
variable to enter into the cumulative sum at the same time.  Test
statistics analogous to those describe above are then computed, with
modified critical values being adopted to reflect the change in the
statistics' computation.

@@

<<theta estimate>>=
# estimate theta 
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by conducting 
log likelihood test, which is done by anova function in R. 
<<LRT>>=
# LRT
anova(theta, thetaAB)
@

\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ 
is a relatively-small number of categories and the sample size 
is not too large (LRT is sensitive to large sample and will 
return "false" significance results, details see next section). 
However, when $V$ (in this case, age) continuous variable, however, 
multiple-group analysis usually cannot be performed because there are 
no existing groups. Instead, we can follow the procedure described in
the Background section.  That is, we can fit a model whose parameters 
are restricted to be equal across across all individuals and then examine 
how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
values of $V$.

%% TO BE MOVED TO BACKGROUND
%% In practice, there are three test statistics can be used: 
%% double maximum, CvM and maxLM. 
%% \begin{equation}
%% DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
%% \end{equation}
%% \begin{equation}
%% CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
%% \end{equation}
%% \begin{equation}
%% maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
%% \end{equation}

To demonstrate these tests in R, we will use the youth gratitude data
and assume that age group is continuous.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we will use the gefp function in strucchange package to 
get the $B(\hat{\theta}_{ij})$. After that, we can get the 
three test statistics' significance by using sctest function.

@@

\readme{Ting: The problem with the last statistic is having
  fplot=TRUE.  To get around this, you can add random noise or set
  fplot=FALSE (I've done the latter).}
<<Get Estimate Alternative Way>>=
DM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="DM", fplot=TRUE) 
DM
CvM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="CvM", fplot=TRUE) 
CvM
maxLM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
maxLM
@

@@

\subsection{Ordinal Test}
However, in this dataset, $V$ is actually ordinal instead of 
continuous. There is only a partial ordering of individuals 
with respect to $V$,i.e. observations with the same level 
of $V$ have no unique ordering. The ordinal statistics 
proposed here are similar to those described in Equation 
and above, except that we focus on "bins" of individuals 
at each level of the ordinal variable. That is, instead 
of aggregating over all $i=1,...n$ individuals, we 
first compute cumulative proportions $t_{l}(l=1,...,m-1)$ 
associated with the first $m-1$ levels of $V$. 
%% TODO Move these details to Background section
We then aggregate the cumulative scores only over 
$i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics 
related can be written as 
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ 
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} 
      ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}

The difference between the continuous test and ordinal 
test is that we don't force arbitary order on the observations 
within the same age groups. Therefore, this ordinal test should 
reflect more of the data. In terms of code, we still need the gefp 
function to get the B estimate. After that, we can use different 
functions to aggregate and scale B estimation. 
The function are called ordwmax and ol2bb1.fun, respectively. 
<<Ordinal Test>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval.ord=rep(NA,2)
pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value

if(file.exists("opval1.rda")){
  load("opval1.rda")
} else {
  opval1 <- ordL2BB(gefp1)
  save(opval1, file="opval1.rda")
}
pval.ord[2]<-sctest(gefp1, functional=opval1)$p.value
pval.ord
@

<<Get Estimate Alternative Way for Ordinal>>=
WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
WDMo 
maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", 
               fnl=opval1, fplot=TRUE)
maxLMo
@

@@

In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. 
Both $p-$values are clearly larger than that of the 
likelihood ratio test and neither is significant at $\alpha=0.05$
